PineAPPL: combining EW and QCD corrections\\ for fast evaluation of LHC processes

### 2020-08-25 paper/paper.tex
Mention PDFs in the abstract; emphasise multi-coupling expansion in abstract and sec.1; discuss choice of processes in sec.3

\abstract{We introduce \textsc{PineAPPL}, a library that produces
  fast-interpolation grids of physical cross sections, computed with a 
  general-purpose Monte Carlo generator, accurate to fixed order in the
  strong, electroweak, and combined strong-electroweak couplings. We
  discuss its features in comparison to similar software
  available in the literature, and we explicitly interface it to
  \textsc{MadGraph5\_aMC@NLO}. We validate the performance of \textsc{PineAPPL}
  by computing fast-interpolation grids, accurate to next-to-leading order in
  the strong and electroweak couplings, for a representative set of LHC
  processes for which EW corrections may have a sizeable effect on the accuracy
  of the corresponding theoretical predictions.
  We formulate a recommendation on the format of the experimental
  deliverables in order to consistently compare them with computations
  that incorporate EW corrections.}

>>>>

\abstract{We introduce \textsc{PineAPPL}, a library that produces
  fast-interpolation grids of physical cross sections, computed with a 
  general-purpose Monte Carlo generator, accurate to fixed order in the
  strong, electroweak, and combined strong-electroweak couplings. We
  demonstrate this unique ability, that distinguishes \textsc{PineAPPL} from
  similar software available in the literature, by interfacing it to
  \textsc{MadGraph5\_aMC@NLO}. We compute fast-interpolation grids,
  accurate to next-to-leading order in the strong and electroweak couplings,
  for a representative set of LHC processes for which EW corrections may have
  a sizeable effect on the accuracy of the corresponding theoretical
  predictions. We formulate a recommendation on the format of the experimental
  deliverables in order to consistently compare them with computations
  that incorporate EW corrections, and specifically to determine
  parton distribution functions to the same accuracy.}

### 2020-08-25 paper/sec-doublecounting.tex
Remove TODO
Mention PDFs in the abstract; emphasise multi-coupling expansion in abstract and sec.1; discuss choice of processes in sec.3

The ability to include theoretical calculations of EW corrections in
PDF determinations is not sufficient to make a \emph{consistent} fit possible. Indeed, experimental data should
be provided in a format that allows PDF collaborations to easily employ them in these fits in a theoretically sound manner. To this purpose, in this section
we formulate some
guidelines to facilitate a consistent presentation of the experimental data. We focus on the problem of data with (partially) subtracted EW effects, which, in PDF fits with theory predictions including them, leads to a double counting issue.

@@

It is interesting
to note that one can tune the QED parton shower to mimic NLO EW effects for specific processes and observables, so that a prediction only accurate at NLO
QCD displays a remarkable agreement with another at NLO QCD+EW when the photonic shower is included (see also the behaviour of predictions showered with
\textsc{Photos}~\cite{Barberio:1990ms,Barberio:1993qi,Golonka:2005pn} in ref.~\cite{CarloniCalame:2016ouw}).
However, this kind of agreement
always comes \emph{a posteriori}, and cannot be ensured in general.
The transverse momentum of the Z boson as shown in section~\ref{sec:Zpt} is such an example: the difference between the two datasets is rather small (less than \SI{6}{\percent}), but the NLO EW corrections are as large as \SI{25}{\percent}.
Furthermore, the deconvolution of QED effects in data introduces a dependence on the program (and possibly on 
the specific version) employed for the shower inversion. 
This fact is especially problematic if deconvolved datasets are the only ones which are published, since undoing the exact deconvolution can be very difficult, or practically impossible.
(\textbf{TODO insert comment about LEP ISR and DIS EW effects removal}).

>>>>

The ability to perform theoretical calculations simultaneously accurate in both
the QCD and EW couplings is not sufficient to make a \emph{consistent}
comparison with experimental measurements. In this section
we formulate some
guidelines to facilitate this task. We focus on the problem of data with (partially) subtracted EW effects, which, if compared to theory predictions including them, leads to a double counting issue.

@@

It is interesting
to note that one can tune the QED parton shower to mimic NLO EW effects for specific processes and observables, so that a prediction only accurate at NLO
QCD displays a remarkable agreement with another at NLO QCD+EW when the photonic shower is included (see also the behaviour of predictions showered with
\textsc{Photos}~\cite{Barberio:1990ms,Barberio:1993qi,Golonka:2005pn} in ref.~\cite{CarloniCalame:2016ouw}).
However, this kind of agreement
always comes \emph{a posteriori}, and cannot be ensured in general.
The transverse momentum of the Z boson as shown in section~\ref{sec:Zpt} is such an example: the difference between the two datasets is rather small (less than \SI{6}{\percent}), but the NLO EW corrections are as large as \SI{25}{\percent}.
Furthermore, the deconvolution of QED effects in data introduces a dependence on the program (and possibly on 
the specific version) employed for the shower inversion. 
This fact is especially problematic if deconvolved datasets are the only ones which are published, since undoing the exact deconvolution can be very difficult, or practically impossible.

### 2020-08-25 paper/sec-introduction.tex
Mention PDFs in the abstract; emphasise multi-coupling expansion in abstract and sec.1; discuss choice of processes in sec.3

The two respects are intertwined, as they both concur to determine the accuracy
of the theoretical predictions that are matched to the precision of the data.
In particular EW (including combined QCD--EW) effects should start to be
incorporated in calculations for LHC processes as standard. The reason is
twofold. First, one expects NNLO QCD and NLO EW corrections to be of
comparable size because, at the EW scale, the QCD and EW running couplings 
become similar, $\alphas^2\sim \alpha$. If NNLO QCD corrections are included
by default in the computations, NLO EW corrections should be
taken into account as well. Second, the virtual exchange of soft or 
collinear weak bosons leads to Sudakov 
logarithms~\cite{Denner:2000jv,Denner:2001gw}
(see also~\cite{Denner:2019vbn} and references therein),
which can make the coefficients of the EW series grow faster than 
their QCD counterparts. This behaviour is relevant in
phase-space regions associated with large mass scales (roughly of the order
of the \si{\tera\electronvolt}), where several LHC data sets (e.g.\ Z-boson
transverse-momentum distributions) enter both the validation of the SM and the search
for new physics.

The systematic inclusion of EW corrections in precision computations for
LHC processes entails
the solution of two separate problems. First, a problem of efficiency:
fast-interpolation grids should be constructed, whereby partonic matrix 
elements, accurate to NLO QCD+EW, are precomputed in such a way that the 
numerical convolution with generic input PDFs can be efficiently approximated
by means of interpolation techniques. Such grids are essential
whenever the evaluation of the hadronic cross section needs to 
be performed a large number of times, as is the case in the evaluation of
scale variations or of PDF fits. While two default formats exist for
these grids, \textsc{APPLgrid}~\cite{Carli:2010rw} and
\textsc{fastNLO}~\cite{Kluge:2006xs,Wobisch:2011ij,Britzger:2012bs}, none of them supports the inclusion of EW
corrections nor the interface to a Monte Carlo generator accurate to 
NLO QCD+EW\@. Second, a problem of consistency: the way in which EW effects may
(or may not) be folded into the data varies across different experimental 
analyses, a fact that challenges their consistent theoretical interpretation. 
Examples are the subtraction of background processes which should not be 
considered as such (e.g.\ the $t$-channel photon-induced component in
neutral-current Drell--Yan, which is not a separate processes beyond leading
order) or of just a part of the EW effects (e.g.\ multiple-photon
radiation from light particles in the final state of neutral- or charged-current
Drell--Yan, especially with electrons). Be that as it may, if EW effects are
systematically included in theoretical predictions, they should not be subtracted
from experimental results, otherwise they will be double counted.

In this paper we address the first of these two problems. Specifically, 
we develop \textsc{PineAPPL}, a library that allows any user to generate
fast-interpolation grids, accurate to up to any fixed order in the strong and
EW couplings. The library supports variations of the factorisation
and renormalisation scales, and can be extended to include
resummation, and matching with a photon- and/or parton-shower.
The grids in the new \textsc{PineAPPL} format complement those
(accurate to fixed-orders in the strong coupling only) that can be generated 
in the \textsc{APPLgrid} and \textsc{FastNLO} formats. The \textsc{PineAPPL}
library is interfaced to \textsc{MadGraph5\_aMC@NLO} (\textsc{mg5\_aMC}
henceforth), with which it has been developed and tested. In this respect,
\textsc{PineAPPL} supersedes
\textsc{APPLgrid}+\textsc{aMCfast}~\cite{Bertone:2014zva}.
\textsc{PineAPPL} can also be easily interfaced to any Monte Carlo generator,
e.g.\ \textsc{SHERPA}~\cite{Biedermann:2017yoi}, where it complements
\textsc{MCgrid}~\cite{DelDebbio:2013kxa,Bothmann:2015dba}.

>>>>

The two respects are intertwined, as they both concur to determine the accuracy
of the theoretical predictions that are matched to the precision of the data.
In particular, taking advantage of the automation pioneered
in~\cite{Kallweit:2014xda,Biedermann:2017yoi,Frederix:2018nkq},
perturbative corrections that arise from the simultaneous expansion in both the
QCD and EW couplings should start to be
incorporated in calculations for LHC processes as standard. The reason is
twofold. First, one expects NNLO QCD and NLO EW corrections to be of
comparable size because, at the EW scale, the QCD and EW running couplings 
become similar, $\alphas^2\sim \alpha$. If NNLO QCD corrections are included
by default in the computations, NLO EW corrections should be
taken into account as well. Second, the virtual exchange of soft or 
collinear weak bosons leads to Sudakov 
logarithms~\cite{Denner:2000jv,Denner:2001gw}
(see also~\cite{Denner:2019vbn} and references therein),
which can make the coefficients of the EW series grow faster than 
their QCD counterparts. This behaviour is relevant in
phase-space regions associated with large mass scales (roughly of the order
of the \si{\tera\electronvolt}), where several LHC data sets (e.g.\ Z-boson
transverse-momentum distributions) enter both the validation of the SM and the search
for new physics.

The consistent inclusion of QCD and EW corrections in precision computations for
LHC processes entails
the solution of two separate problems. First, a problem of efficiency:
fast-interpolation grids should be constructed, whereby partonic matrix 
elements, accurate to NLO QCD+EW, are precomputed in such a way that the 
numerical convolution with generic input PDFs can be efficiently approximated
by means of interpolation techniques. Such grids are essential
whenever the evaluation of the hadronic cross section needs to 
be performed a large number of times, as is the case in the evaluation of
scale variations or of PDF fits. While two default formats exist for
these grids, \textsc{APPLgrid}~\cite{Carli:2010rw} and
\textsc{fastNLO}~\cite{Kluge:2006xs,Wobisch:2011ij,Britzger:2012bs}, none of them supports the inclusion of EW
corrections nor the interface to a Monte Carlo generator accurate to 
NLO QCD+EW\@. Second, a problem of consistency: the way in which EW effects may
(or may not) be folded into the data varies across different experimental 
analyses, a fact that challenges their consistent theoretical interpretation. 
Examples are the subtraction of background processes which should not be 
considered as such (e.g.\ the $t$-channel photon-induced component in
neutral-current Drell--Yan, which is not a separate processes beyond leading
order) or of just a part of the EW effects (e.g.\ multiple-photon
radiation from light particles in the final state of neutral- or charged-current
Drell--Yan, especially with electrons). Be that as it may, if EW effects are
systematically included in theoretical predictions, they should not be subtracted
from experimental results, otherwise they will be double counted.

In this paper we address the first of these two problems. Specifically, 
we develop \textsc{PineAPPL}, a library that allows any user to generate
fast-interpolation grids, accurate to up to any fixed order in the QCD and
EW couplings. The library supports variations of the factorisation
and renormalisation scales, and can be extended to include
resummation, and matching with a photon- and/or parton-shower.
The grids in the new \textsc{PineAPPL} format complement those
(accurate to fixed-orders in the strong coupling only) that can be generated 
in the \textsc{APPLgrid} and \textsc{FastNLO} formats. The \textsc{PineAPPL}
library is interfaced to \textsc{MadGraph5\_aMC@NLO} (\textsc{mg5\_aMC}
henceforth), with which it has been developed and tested. In this respect,
\textsc{PineAPPL} supersedes
\textsc{APPLgrid}+\textsc{aMCfast}~\cite{Bertone:2014zva}.
\textsc{PineAPPL} can also be easily interfaced to any Monte Carlo generator,
e.g.\ \textsc{SHERPA}~\cite{Biedermann:2017yoi}, where it complements
\textsc{MCgrid}~\cite{DelDebbio:2013kxa,Bothmann:2015dba}.

### 2020-08-25 paper/sec-results.tex
Mention PDFs in the abstract; emphasise multi-coupling expansion in abstract and sec.1; discuss choice of processes in sec.3

In this section we demonstrate the capabilities of \textsc{mg5\_aMC}+\textsc{PineAPPL} by
computing fast-interpolation grids, accurate to NLO QCD and NLO QCD+EW,
for a representative set of processes in which EW corrections may have a
sizeable effect on the accuracy of the theoretical predictions.
In order to consider some realistic kinematics for these
processes, we rely on a set of measurements that are commonly included in PDF
fits. Our aim is twofold. First, we want to validate the results
obtained with \textsc{PineAPPL}; second, we want to assess the
impact of the EW corrections for the specific experimental setups. We describe
the settings employed for the computations, and the corresponding results for
each process. We remind the reader that \textsc{mg5\_aMC} makes it possible
to compute predictions including NLO QCD and EW corrections for arbitrary processes
in an automated manner. It employs the FKS subtraction scheme~\cite{Frixione:1995ms,Frixione:1997np}
as automated in \textsc{MadFKS}~\cite{Frederix:2009yq,Frederix:2016rdc} to deal with IR singularities. One-loop
amplitudes are computed by \textsc{MadLoop}~\cite{Hirschi:2011pa}, which employs different
numerical techniques~\cite{Passarino:1978jh,Davydychev:1991va,Denner:2005nn,Ossola:2006us,Cascioli:2011va,Mastrolia:2012bu} implemented in the corresponding computer libraries~\cite{Ossola:2007ax,Peraro:2014cba,Hirschi:2016mdz,Denner:2016kdg}. Matching
with parton showers is available only for pure-QCD corrections via the MC@NLO method~\cite{Frixione:2002ik}, and 
will not be employed in the following.

@@

@@

We focus on the following three processes: Drell--Yan lepton-pair production,
top-quark pair production, and Z-boson (lepton-pair) production with non-zero
transverse momentum in proton-proton collisions. In the following, we shall
present the experimental measurements, the process-specific settings, and
the phenomenological results for each of these processes.

>>>>

In this section we demonstrate the capabilities of the \textsc{PineAPPL}
library by interfacing it to \textsc{mg5\_aMC}, and by
computing fast-interpolation grids, accurate to NLO QCD and NLO QCD+EW,
for a common set of LHC processes in which EW corrections may sizeably
affect the accuracy of the theoretical predictions.
In order to consider some realistic kinematics for these
processes, we rely on a representative set of LHC measurements. Our aim is twofold. First, we want to validate the results
obtained with \textsc{PineAPPL}; second, we want to assess the
impact of the EW corrections for the specific experimental setups. We describe
the settings employed for the computations, the corresponding results for
each process, and possible implications for the determination of PDFs.

@@

\textsc{mg5\_aMC} makes it possible
to compute predictions including NLO QCD and EW corrections for arbitrary processes
in an automated manner. It employs the FKS subtraction scheme~\cite{Frixione:1995ms,Frixione:1997np}
as automated in \textsc{MadFKS}~\cite{Frederix:2009yq,Frederix:2016rdc} to deal with IR singularities. One-loop
amplitudes are computed by \textsc{MadLoop}~\cite{Hirschi:2011pa}, which employs different
numerical techniques~\cite{Passarino:1978jh,Davydychev:1991va,Denner:2005nn,Ossola:2006us,Cascioli:2011va,Mastrolia:2012bu} implemented in the corresponding computer libraries~\cite{Ossola:2007ax,Peraro:2014cba,Hirschi:2016mdz,Denner:2016kdg}. Matching
with parton showers is available only for pure-QCD corrections via the MC@NLO method~\cite{Frixione:2002ik}, and 
will not be employed in the following.

@@

We focus on the following three processes: Drell--Yan lepton-pair production,
top-quark pair production, and Z-boson (lepton-pair) production with non-zero
transverse momentum in proton-proton collisions. These are some of the most
commonly and most precisely measured processes at the LHC, which are widely
studied to test the SM and/or search for new physics. We therefore expect them
to allow us to clearly show the benefit of being able to make fast and
reliable theoretical predictions accurate to NLO QCD+EW with \textsc{PineAPPL}.
In the following, we shall
present the experimental measurements, the process-specific settings, and
the phenomenological results for each of these processes.
